---
output:
  github_document
---

# SMiXcan: Cell-type-aware TWAS from bulk transcriptomics using GWAS summary statistics

**SMiXcan** is a summary-statistics-based, cell-type-aware TWAS framework for studying how genetically predicted expression relates to disease risk across multiple cell types.

This document walks through the basic SMiXcan workflow using the small example datasets bundled with the package.
The goal is to show the expected inputs and outputs for each function, not to support biological interpretation.

------------------------------------------------------------------------

## Overview of the workflow

The SMiXcan pipeline consists of four modular steps:

1. Estimate cell-type fractions from bulk expression
2. Train cell-type–specific gene expression prediction models
3. Perform SMiXcan association testing
4. (Optional) Classify cell-type association patterns

Each step corresponds to one exported function.

------------------------------------------------------------------------

## Installation

With R, users can install the SMiXcan package directly from GitHub with [devtools](https://github.com/hadley/devtools):

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("songxiaoyu/SMiXcan")
```

------------------------------------------------------------------------

## Step 1: Estimate cell-type fractions (Optional)

Cell-type fractions represent the proportion of each cell type contributing to each bulk RNA-seq sample. This step is optional because users may already have their own estimated cell-type proportions.

### Parameters

- `exprB`
  Bulk expression matrix (genes × samples).

- `markers`
  Named list of marker genes defining each cell type. The list length determines the number of cell types K.

- `seed`
  Random seed for reproducibility.

- `n.iter`, `burn.in`
  MCMC settings passed to the underlying estimator.

```{r step1-pi}
library(SMiXcan)
data(exprB_example, markers_example)
head(exprB_example)
head(markers_example)
res_pi <- pi_estimation_K(
  exprB = exprB_example,
  markers = markers_example,
  seed = 1,
  n.iter = 1000,
  burn.in = 200
)
head(res_pi$cell_fraction)
```

### Output

- `cell_fraction`
  A data frame with one row per sample and one column per cell type.

------------------------------------------------------------------------

## Step 2: Train cell-type–specific MiXcan models

Next, we train genetic prediction models that allow SNP effects on expression to vary across cell types using a symmetric parameterization.

### Parameters

- `y`
  Expression vector for a single gene (length = number of samples).

- `x`
  Genotype matrix for cis-SNPs (samples × SNPs).

- `pi_k`
  Cell-type fraction matrix (samples × K).

- `yName`
  Optional gene identifier.

```{r step2-training}
data(x_example, y_example, pi_k)

set.seed(1)
foldid <- sample(rep(1:4, length.out = nrow(x_example)))

fit <- MiXcan_train_K(
  y      = y_example,
  x      = x_example,
  pi_k   = pi_k,
  foldid = foldid,
  yName  = "GENE_A1"
)

fit$type
fit$W
```

### Output of `MiXcan_train_K()`

The object `fit` returned by `MiXcan_train_K()` is a list containing:

- `type`
  Model classification: `"CellTypeSpecific"`, `"NonSpecific"`, or `"NoPredictor"`.

- `beta.SNP.by.cell`
  SNP weights returned separately for each cell type.

- `beta.all.models`
  Coefficients from the tissue-level model and the cell-type-specific models.

- `W`
  Cell-type-specific SNP weight matrix used in downstream association testing.

- `glmnet.cell`
  Fitted penalized model for the K-cell-type regression.

- `glmnet.tissue`
  Fitted tissue-level baseline model.

- `yName`
  Returned gene label.

- `xNameMatrix`
  Returned SNP annotation.

------------------------------------------------------------------------

## Step 3: SMiXcan association testing

SMiXcan tests gene–trait associations by combining trained SNP weights with GWAS summary statistics while accounting for LD.

### Parameters

- `W`
  Cell-type-specific SNP weight matrix from `MiXcan_train_K()`.

- `gwas_results`
  GWAS summary statistics with `Beta` and `se_Beta`.

- `x_g`
  Reference-panel genotype matrix.

- `n0`, `n1`
  Numbers of controls and cases.

- `family`
  Outcome type, either `"binomial"` or `"gaussian"`.

```{r step3-association}
W <- fit$W

res_assoc <- SMiXcan_assoc_test_K(
  W            = W,
  gwas_results = gwas_example,
  x_g          = x_example,
  n0           = 1000,
  n1           = 1000,
  family       = "binomial"
)

res_assoc
```


### Output of `SMiXcan_assoc_test_K()`

- `Z_join`
  Joint Z-scores for each cell type.
- `p_join_vec`
  Joint p-values for each cell type.
- `p_join`
  Combined ACAT p-value across cell types.

### Interpretation

- If `p_join` is significant, the gene shows evidence of association.
- `p_join_vec` indicates which cell types contribute most strongly.
- The sign of `Z_join` indicates direction of effect.


------------------------------------------------------------------------

## Step 4: PRIMO-based cell-type association pattern classification (optional)

After running SMiXcan genome-wide, we may wish to classify significant
genes into cell-type association patterns using a PRIMO-based framework.

This step:

1. Adjusts marginal and joint p-values using BH-FDR  
2. Identifies significant genes  
3. Estimates PRIMO posterior probabilities  
4. Assigns MAP patterns among non-null configurations  

```{r step4-primo, warning=FALSE, message=FALSE}
data(merged_example)

head(merged_example)

res_primo <- infer_celltype_patterns(
  merged          = merged_example,
  pvals_names     = c("p_1_ct2", "p_2_ct2"),
  p_join_name     = "p_join_ct2",
  type_col        = "type_ct2",
  fdr_cutoff      = 0.1
)

length(res_primo$genes_sig_specific)
length(res_primo$genes_sig_nonspecific)
lengths(res_primo$unique_by_cell)
res_primo$tab_specific_patterns
```

### Parameters

- `merged`  
  Data frame containing marginal p-values, a joint p-value, and a cell-type-specificity label.

- `pvals_names`  
  Character vector of marginal p-value column names (length = K).

- `p_join_name`  
  Column name of joint (non-specific) p-value.

- `type_col`  
  Column indicating `"CellTypeSpecific"` vs `"NonSpecific"` genes.

- `fdr_cutoff`  
  FDR threshold used to define significance (default 0.1).

---

### Key Outputs

- `out`  
  Input table augmented with BH-adjusted p-values, PRIMO posterior probabilities (`post_*`), and MAP pattern labels (`MAP_pattern_nonnull`).

- `genes_sig_specific`  
  Gene IDs for significant cell-type-specific genes.

- `genes_sig_nonspecific`  
  Gene IDs for significant non-specific genes based on the joint test.

- `unique_by_cell`  
  A named list of genes uniquely assigned to each cell type pattern.

- `shared_specific_genes`  
  Significant cell-type-specific genes assigned to patterns involving two or more cell types.

- `tab_specific_patterns`  
  Table summarizing MAP pattern counts among significant cell-type-specific genes.


### Interpretation

For K = 2 cell types, PRIMO considers the following patterns:

- `"10"` : associated only with cell type 1  
- `"01"` : associated only with cell type 2  
- `"11"` : associated with both cell types  

MAP classification is performed **only among significant cell-type-specific genes**
and excludes the null pattern (`"00"`).

------------------------------------------------------------------------
